ExpressionView -- an interactive viewer for modules identified in gene expression data

### 2010-01-15 expressionviewnote/expressionview.tex
EV note updates, should be ready to submit.

\firstpage{1}
\application{}
\title[ExpressionView]{ExpressionView - an interactive viewer for
  modules identified in gene expression data}
\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de
Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
  L\"uscher$^{1*}$, G\'abor Cs\'ardi$^{1,2*}$, Aitana Morton de
  Lachapelle$^{1,2*}$, Zolt\'an Kutalik$^{1,2*}$,
  Bastian Peter$^{1,2}$
  and Sven Bergmann$^{1,2}$}
\address{
  $^{1}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\
  $^{2}$Department of Medical Genetics, University of Lausanne,
  Lausanne, Switzerland\\
  *: equal contribution
}

@@

\vspace*{-7pt}
\section{Introduction}
Biclustering is an unsupervised data analysis method which is
frequently used to explore microarray data. Biclustering algorithms
process collections of expression profiles to identify co-expressed groups of
genes and conditions (samples) for which this co-expression occurs. 
We refer to such groups as modules. While
there is a multitude of biclustering software available (for a reviews 
and comparisons see~\cite{madeira04,ihmels04,prelic06}), packages with 
intuitive interfaces that allow for an interactive exploration of the results are
sparse~\citep{santamaria08}. ExpressionView is designed to close this
gap, and facilitate the collaboration between bioinformaticians and
life scientists without consolidated programming
experience. Implemented as an R package, the user can apply all the
powerful microarray analysis tools provided by the
Bioconductor~\citep{gentleman04} project before exporting the data to
a platform independent Adobe Flash applet. This applet
provides an integrated, intuitive and interactive visualization of
the modules together with the underlying gene expression data and
additional information on the genes and conditions.

@@

\subsection{Reordering genes and conditions}
To present the collection of possibly overlapping modules in a visually
appealing form, it is necessary to reorder the rows (conditions) and
columns (genes) of the gene expression matrix in such a way that biclusters form contiguous
rectangles. Since it is in general impossible to find such an
arrangement for more than two mutually overlapping modules, we propose
here an approximate solution that optimizes the arrangement within the
original data by maximising the total area of the largest contiguous
module subsets. (An alternative would be to repeat rows and columns as
necessary~\citep{grothaus06}, but for many modules this results in
a very large expression matrix.)

This optimization task is an interesting problem on its own, which to
the best of our knowledge has not been studied in the literature. Thus
we briefly outline our strategy here (see our website for details).
The reordering of the rows and the columns do not depend on each
other, so the same optimization method can be applied independently
to rows and columns. For a given order of the
elements (either genes or conditions) we compute for each module $i$ the
size of the largest contiguous sequence of elements (i.e. the maximal
number of neighboring elements $N^\text{max}_i$). Then, as a measure of the
quality ($Q$) of the order we sum this quantity over all modules
($Q=\sum_i N^\text{max}_i$). To optimize $Q$, an initial sequence is calculated using
hierarchical clustering. Two operations are then applied to this: (1)
Permutations that exchange two elements within a module and (2) shifts
of a sequence of multiple elements of the same module to a different position. We
use a greedy iterative scheme that performs these operations at
randomly selected positions and keeps the new sequence if it improves
$Q$. The algorithm stops if after a given number of operations no
significant improvement of $Q$ is achieved.

>>>>

\firstpage{1}
\application{}
\title[ExpressionView]{ExpressionView -- an interactive viewer for
  modules identified in gene expression data}
\author[Andreas L\"uscher, G\'abor Cs\'ardi, Aitana Morton de
Lachapelle, Zolt\'an Kutalik, and Sven Bergmann]{Andreas
  L\"uscher$^{1*}$, G\'abor Cs\'ardi$^{1,2*}$, Aitana Morton de
  Lachapelle$^{1,2*}$, Zolt\'an Kutalik$^{1,2*}$,
  Bastian Peter$^{1,2}$
  and Sven Bergmann$^{1,2}$}
\address{
  $^{1}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\
  $^{2}$Department of Medical Genetics, University of Lausanne,
  Lausanne, Switzerland\\
  *: equal contribution
}

@@

\vspace*{-7pt}
\section{Introduction}
Biclustering is an unsupervised data analysis method which is
frequently used to explore microarray data. 
Biclustering algorithms process collections of expression profiles to
identify groups of genes co-expressed under some conditions (samples).
We refer to such groups as modules. While
there is a multitude of biclustering software available (for a reviews 
and comparisons see~\cite{madeira04,ihmels04,prelic06}), packages with 
intuitive interfaces that allow for an interactive exploration of the results are
sparse~\citep{santamaria08}. ExpressionView is designed to close this
gap, and facilitate the collaboration between bioinformaticians and
life scientists without consolidated programming
experience. Implemented as an R package, the user can apply all the
powerful microarray analysis tools provided by the
Bioconductor project~\citep{gentleman04} before exporting the data to
a platform independent Adobe Flash applet. This applet
provides an integrated, intuitive and interactive visualization of
the modules together with the underlying gene expression data and
additional information on the genes and conditions.

@@

\subsection{Reordering genes and conditions}
To present the collection of possibly overlapping modules in a visually
appealing form, it is necessary to reorder the rows (conditions) and
columns (genes) of the gene expression matrix in such a way that biclusters form contiguous
rectangles. Since it is in general impossible to find such an
arrangement for more than two mutually overlapping modules, we propose
here an approximate solution that optimizes the arrangement within the
original data, by maximising the total area of the largest contiguous
module subsets. (An alternative would be to repeat rows and columns as
necessary~\citep{grothaus06}, but for many modules this results in
a very large expression matrix.)

This optimization task is an interesting problem on its own, which to
the best of our knowledge has not been studied in the literature. Thus
we briefly outline our strategy here (see our website for details).
The reordering of the rows and the columns do not depend on each
other, so the same optimization method can be applied independently
to rows and columns. For a given order of the
elements (either genes or conditions) we compute for each module $i$ the
size of the largest contiguous sequence of elements (i.e. the maximal
number of neighboring elements $N^\text{max}_i$). Then, as a measure of the
quality ($Q$) of the order we sum this quantity over all modules
($Q=\sum_i N^\text{max}_i$). To optimize $Q$, an initial sequence is calculated using
hierarchical clustering. Two operations are then applied to this: (1)
permutations that exchange two elements within a module and (2) shifts
of a sequence of multiple elements of the same module to a different position. We
use a greedy iterative scheme that performs these operations at
all possible positions and keeps the best new sequence if it improves
$Q$. The algorithm stops if after a given number of operations no
significant improvement of $Q$ is achieved.
